Virtualization of a Real-Time Operating System for Robot Control with a Focus on Real-Time Compliance

### 2024-08-31 Masterarbeit_Robotik_Pamuk.tex
Correct grammar

\hypersetup{
    colorlinks=true, % Aktiviert farbige Links
    linkcolor={rgb,255:red,0;green,0;blue,255}, % Interne Links (z.B. Sections)
    citecolor=orange, % Zitate
    urlcolor={rgb,255:red,128;green,0;blue,128} % URLs
}

@@

\bigskip \noindent Erste Latenzmessungen zeigten eine signifikante Leistungslücke zwischen der Bare-Metal und dem virtualisierten Setup. Daher wird ein umfangreicher Tuning-Prozess durchgeführt, um Echtzeit-Performance und Determinismus zu erreichen. Diese Modifikationen umfassen Konfigurationen, die das BIOS, den Kernel, das Host-Betriebssystem, die QEMU/KVM-Virtualisierungsschicht und das Salamander 4-Betriebssystem (Gast) selbst betreffen. Die Worst-Case-Latenz wurde von 707.622\textmu s auf 17.134\textmu s reduziert, was der Bare-Metal-Performance von 10.709\textmu s nahekommt. Darüber hinaus wird die Verbesserung der Echtzeit-Performance und des Determinismus anhand einer Roboteranwendung validiert, bei der die abgestimmte Virtualisierung mit der nicht abgestimmten und der Bare-Metal-Version verglichen wird.

\bigskip \noindent Insgesamt bietet diese Arbeit eine umfassende Blaupause, um ein virtualisiertes Gastsystem in einem Host-System mit deterministischem Verhalten echtzeitfähig zu machen.}
\schlagworte{Virtualisierung, Echtzeitsysteme, Latenzreduktion, Robotiksteuerung}

@@

\bigskip \noindent Initial latency measurements revealed a significant performance gap between the bare metal and virtualized setup. Thus, an extensive tuning process is carried out to achieve real-time performance and determinism. These modifications involve configurations spanning the BIOS, kernel, host operating system, QEMU/KVM virtualization layer, and the Salamander 4 operating system (guest) itself. The worst-case latency was brought down from 707.622\textmu s to 17.134\textmu s, closely aligning with the bare metal performance of 10.709\textmu s. In addition, the improvement in real-time performance and determinism is validated using a robotic application, where the tuned virtualization is compared with the untuned and the bare metal version.

@@

\noindent However, in an RTOS, a high-priority process can interrupt a lower-priority task, even if it is in the middle of a kernel operation, as shown in Figure~\ref{fig:kernel_rt} RTOS are specifically designed to react to events within fixed time limits and prioritise the execution of high-priority processes.

@@

\section{Real-Time Operating Systems}
\noindent The RTOS structure is visualized in Figure~\ref{fig:rtos_structure}.The core component of an RTOS that enables real-time capabilities is the kernel. It is responsible for managing system resources, scheduling tasks, and ensuring deterministic behavior~\cite{malallahComprehensiveStudyKernel2021}. It employs preemptive scheduling mechanisms to allow high-priority tasks to preempt lower-priority tasks, ensuring that time-critical tasks are not delayed. Additionally, RTOS kernels are designed to allocate and manage memory resources efficiently and minimize interrupt latency, which is crucial for real-time applications that require immediate response to external events~\cite{wangRealtimeEmbeddedSystems2017}.

@@

\noindent In these real-time operating system, task scheduling is based on so-called priority-based preemptive scheduling~\cite{buttazzoHardRealtimeComputing2024}. Each task in a software application is assigned a priority. A higher priority means that a faster response is required. Preemptive task scheduling ensures a very fast response. Preemptive means that the scheduler can stop a currently running task at any point if it recognizes that another task needs to be executed immediately. The basic rule on which priority-based preemptive scheduling is based is that the task with the highest priority that is ready to run is always the task that must be executed. So if both a task with a lower priority and a task with a higher priority are ready to run, the scheduler ensures that the task with the higher priority runs first. The lower priority task is only executed once the higher priority task has been processed.

@@

\begin{itemize}
	\item This work was written at SIGMATEK GmbH \& Co KG, a company that researches, develops, produces, and distributes automation technology for industrial machinery and plant engineering. SIGMATEK uses its own custom Linux-based operating system, to be run on their self-manufactured CPUs. This operating system will be referred to as ``Salamander 4'' in this work. The details of Salamander 4 are explained in Chapter~\ref{sec:guest_operating_system}.
	\item Salamander 4 was created with Yocto~\cite{WelcomeYoctoProject}, an open-source project that helps develop custom Linux-based systems. Yocto is discussed in Section~\ref{sec:yocto}.
	\item Salamander 4 is virtualized through Quick emulator (QEMU), an open-source emulator and virtualizer that allows to run different operating systems on a computer QEMU is described in Section~\ref{sec:qemu}.
	\item Salamander 4 employs hard real-time with Xenomai 3~\cite{XenomaiXenomai}, a real-time framework for Linux that enables deterministic and low-latency performance. Xenomai 3 is detailed in Section~\ref{sec:xenomai}.
	\item The goal is to virtualize Salamander 4 and approach the performance of bare metal (hardware) through real-time performance tunings, focussing on latency requirements set by SIGMATEK. These modifications are subject of Chapter~\ref{sec:real-time_tunings}.
	\item The \texttt{latency} tool of the Xenomai suite to measure the scheduling latency of the real-time OS was suggested to use by SIGMATEK. The goal was set to achieving latency values below 50 microseconds in the duration of the measuremenet.
	\item SIGMATEK uses the VARAN bus for communication between machines and systems, a real-time Ethernet network that was specially developed for industrial automation. Its operating principle is outlined in Section~\ref{sec:varan}.
	\item For the purpose of replacing the physical CPU, SIGMATEK provided a PCI Insert Card module PCV 522, that was plugged into the PC, as explained in Subsection~\ref{subsec:setup_experiment_virtualized}. This component could then be used to communicate with the Input/Output peripherals, just like the physical CPU could when the I/O modules were attached to it. Subsequently, the PCV 522 module of SIGMATEK was connected with the Pulse Width Module PW 022 of SIGMATEK over the VARAN bus and the VI 021 module of SIGMATEK serving as the power supply.
	\item The robotic application from Chapter~\ref{sec:robotic_application} was written in Lasal Class 2, an object oriented programming tool of SIGMATEK and part of SIGMATEK`s Engineering Platform LASAL, with client-server technology and graphic representation.
\end{itemize}
\section{Related Work and State of the Art}
This master’s thesis builds upon a number of previous studies in the field of virtualization of real-time systems and the inherent latency. In this section, the most influential studies are presented and their relevance to the work is addressed.

\bigskip \noindent \citeauthor{perneelRealtimeCapabilitiesStandard2015}~\cite{perneelRealtimeCapabilitiesStandard2015} mention in their work that Linux was initially developed as a general purpose operating system without real-time applications in mind. They talk about the evolution of real-time behavior in the standard Linux kernel and state that it has recently gained popularity among the real-time community, largely due to its open-source nature and stability. The authors explain how to transform Linux into a real-time operating system by implementing the PREEMPT-RT patches. The paper also emphasizes some kernel configurations that ensure that the system's real-time behavior is preserved during runtime. Even though these modifications to the kernel enable soft real-time performance, the authors underline the fundamental rule in real-time software, that latency improvements have a negative impact on throughput and kernel performance. This is due to the added overhead of the \texttt{CONFIG\_PREEMPT\_RT} option. They concluded with their Linux build and a testing application that Linux can be a viable option for real-time applications, if it is correctly configured. \citeauthor{reghenzaniRealTimeLinuxKernel2020}~\cite{reghenzaniRealTimeLinuxKernel2020} provide an extensive overview on the real-time Linux kernel research, specifically on the PREEMPT RT patch. \citeauthor{adamRealTimePerformanceResponse2021}~\cite{adamRealTimePerformanceResponse2021} applied the PREEMPT\_RT patch on both Raspberry Pi 3 and BeagleBone Black and achieved significantly lower latencies compared to standard Linux kernels.

\bigskip \noindent \citeauthor{cinqueVirtualizingMixedCriticalitySystems2022}~\cite{cinqueVirtualizingMixedCriticalitySystems2022} survey different virtualization approaches, including separation kernels, microkernels, enhanced general-purpose hypervisors, and lightweight virtualization solutions like containers and unikernels. Similarly, \citeauthor{sandstromVirtualizationTechnologiesEmbedded2013}~\cite{sandstromVirtualizationTechnologiesEmbedded2013}, and~\citeauthor{taccariEmbeddedRealTimeVirtualization2014}~\cite{taccariEmbeddedRealTimeVirtualization2014} discuss and compare the current state-of-the-art in virtualization technologies with a focus on embedded real-time platforms. The latter emphasizes, that research on real-time systems is focusing on virtualization and multicore scheduling and underlines that flexibility and cost are key metrics. Both works acknowledge the great benefits of virtualization, that is, reducing overall hardware costs, since all non-critical activities and real-time control tasks can run on the same hardware. In addition,~\citeauthor{javierperezHowRealTime2022}~\cite{javierperezHowRealTime2022} state that hardware-based Programmable Logic Controllers (PLCs) are costly to maintain and lack the flexibility needed for modern, resource-intensive applications like Machine Learning (ML) or Artificial Intelligence (AI).

@@

\bigskip \noindent \citeauthor{maPerformanceTuningKVMbased2013}~\cite{maPerformanceTuningKVMbased2013}~\cite{junzhangPerformanceAnalysisKVMBased2010} describe in their work how KVM impacts the real-time performance, which has helped this work gain insight into the base overhead of virtualization. When a physical interrupt occurs while the guest RTOS is running, there are at least six stages that need to be traversed before the interrupt can be delivered to the guest RTOS. Once the interrupt is delivered to the guest RTOS, it incurs additional latency, called Interrupt Routine Time (IRT), since there is also the time between when an interrupt is recognized and the first instruction of the corresponding interrupt service routine (ISR) of the guest OS is started. These stages are the primary sources of latencies caused by KVM, they are briefly described in Figure~\ref{fig:kvm_stages}.
\vspace{2em}
\begin{figure}[H]
	\centering
	\begin{tikzpicture}[node distance = 0.35cm, auto, % Adjusted node distance
			block/.style={
					rectangle,
					draw,
					fill=blue!20,
					text width=30em,
					rounded corners,
					align=center,
					inner sep=0.5em
				},
			line/.style={draw, -latex'}
		]
		% Place nodes
		\node [block] (init) {\scriptsize \textbf{Guest Code}\\  The guest OS is executing its normal operations.};
		\node [block, below=of init] (vmexit) {\scriptsize \textbf{VM Exit}\\  External interrupt triggers VM exit. Hardware stores interrupt info in \\\vspace{-0.5em} Virtual Machine Control Structure (VMCS).};
		\node [block, below=of vmexit] (aftervmexit) {\scriptsize \textbf{After VM Exit}\\  Control returns to kernel space. KVM retrieves interrupt info from VMCS,\\\vspace{-0.5em} marks it as pending for guest.};
		\node [block, below=of aftervmexit] (hostisr) {\scriptsize \textbf{Host ISR}\\  Host Linux handles the interrupt.};
		\node [block, below=of hostisr] (hostscheduling) {\scriptsize \textbf{Host Scheduling}\\  Host kernel invokes scheduler. Ideally, guest RTOS process is not preempted.};
		\node [block, below=of hostscheduling] (beforevmentry) {\scriptsize \textbf{Before VM Entry}\\  KVM's kernel module prepares for VM entry. Detects pending interrupt,\\\vspace{-0.5em} injects virtual interrupt into guest by writing to VMCS.};
		\node [block, below=of beforevmentry] (vmentry) {\scriptsize \textbf{VM Entry}\\  Hardware emulates conventional interrupt handling for guest based on VMCS info.};
		\node [block, below=of vmentry] (guestirt) {\scriptsize \textbf{Guest IRT}\\  Guest's interrupt handler is invoked to service the interrupt.};
		% Draw edges
		\path [line] (init) -- (vmexit);
		\path [line] (vmexit) -- (aftervmexit);
		\path [line] (aftervmexit) -- (hostisr);
		\path [line] (hostisr) -- (hostscheduling);
		\path [line] (hostscheduling) -- (beforevmentry);
		\path [line] (beforevmentry) -- (vmentry);
		\path [line] (vmentry) -- (guestirt);
	\end{tikzpicture}
	\caption[Flowchart of Operations in a virtualized Environment]{Flowchart of Operations in a virtualized Environment (Own Figure, inspired by~\citeauthor{maPerformanceTuningKVMbased2013})}
	\label{fig:kvm_stages}
\end{figure}
\vspace{1em}

@@

\bigskip \noindent \citeauthor{broskyShieldedProcessorsGuaranteeing2003}~\cite{broskyShieldedProcessorsGuaranteeing2003} report in their work about the shielded CPU model for enhancing real-time performance in symmetric multiprocessor systems. By dedicating one or more CPUs to high-priority processes and interrupts, the goal is to provide deterministic execution of real-time applications and interrupt responses. The authors mention that there is no user interface for setting process CPU affinity in the standard Linux, while there is one for setting interrupt CPU affinity via the \texttt{/proc/irq/*/smp\_affinity} files. These files were used in this masther's thesis. According to the authors, the shielded CPU model is especially useful for tasks that require guaranteed interrupt response times, very fast interrupt responses and high-frequency, deterministic execution.

@@

\bigskip \noindent \citeauthor{yoonRealTimePerformanceAnalysis2009}~\cite{yoonRealTimePerformanceAnalysis2009} use some of these real-time tunings in their research, including CPU shielding, memory locking, and spinning nanosleep, for controlling humanoid robots equipped with around 60 servo motors and sensors. They implement EtherCAT for real-time communication of distributed devices whereas this thesis uses Varan. The authors highlight the importance of timely data processing in robots that collect data from their environment and respond accordingly through their sensors and actuators. The authors conclude with an experiment that their robotic system could achieve deterministic control of actuators and sensors even under heavy system load through the use of Linux with the real-time preemption patch. Their research has similarities to this master's thesis in terms of real-time tunings for robot control. However, this thesis includes also the virtualization aspect that is central.

\bigskip \noindent \citeauthor{kiszkaLinuxRealTimeHypervisor2009}~\cite{kiszkaLinuxRealTimeHypervisor2009} specifically focuses on Linux as a hypervisor and analyzes its real-time capabilities when using KVM and QEMU. Most importantly for this master's thesis, the author warns of starvation when the priorities of QEMU’s threads are raised and lifted into a real-time scheduling class. Starvation is when a process does not get the resources it needs for a long time because the resources are being allocated to other processes. A way to reduce this risk is by not using more virtual CPUs (vCPUs) than there are actual processor cores. Moreover, Linux has a feature that can limit how much CPU time all real-time tasks can use in the host system, which was an important information for the real-time tunings of this thesis later.

@@

\bigskip \noindent \citeauthor{mckenneyRealTimeVs2008}~\cite{mckenneyRealTimeVs2008} addresses the overheads associated with real-time Linux when using KVM and QEMU. The author lists several sources of overhead, including memory locking to avoid page-fault latencies, increased overhead of locking and interrupts due to more-aggressive preemption, threaded interrupts that permit long-running interrupt handlers to be preempted by high-priority real-time processes, real-time task scheduling that requires global scheduling, high-resolution timers for tens-of-microseconds accuracy and precision, and preemptible RCU which has slightly higher read-side overhead than Classic RCU. These information helped understand the trade-offs involved in real-time Linux systems.

\bigskip \noindent \citeauthor{deoliveiraDemystifyingRealTimeLinux2020}~\cite{deoliveiraDemystifyingRealTimeLinux2020} critique the traditional \texttt{cyclictest} tool used for measuring scheduling latency, highlighting its limitations due to its black-box approach and lack of theoretical grounding. This is the main reason why this tool was not used to measure the scheduling latency in this work. Instead, the \texttt{latency} tool was used, as was determined in Chapter~\ref{sec:application_context}.

\section{Problem and Task Definition}
As stated previously, virtualization of real-time systems has many advantages in comparison to traditional hardware-based approaches. Solutions based on pure hardware are often built for a distinct purpose and are hence limited in flexibility. It is difficult to adapt the hardware topology to new requirements or new applications. Moreover, during periods when the demand is lower, expensive hardware may sit idle, leading to inefficient use of resources and a poor return on investment~\cite{gabrielResourceAwareParameterTuning2019}. Another crucial factor is of course the high cost factor associated with procuring and maintaining separate hardware for each real-time system~\cite{bhardwajVirtualizationCloudComputing2021}. Especially when organizations scale their operations, they have to add more physical machines, and the expenses rise rapidly~\cite{abbasiExploringOpenStackScalable2023}. Besides, one needs physical access to the hardware for any updates or maintenance, which is hard to achieve, especially for systems in remote or difficult-to-access locations. This can lead to system downtime and may not only disrupt the operations but can also result in lost productivity and potential revenue~\cite{mahfoudRealTimePredictiveMaintenanceBased2024}.

\bigskip \noindent Virtualization promises to overcome these challenges~\cite{queirozContainerbasedVirtualizationRealtime2024}~\cite{bhardwajVirtualizationCloudComputing2021}~\cite{cinqueEvaluatingVirtualizationFog2023}. However, virtualization also has its own challenges and pitfalls. The additional layers of abstraction can lead to increased overhead and latency~\cite{casiniLatencyAnalysisVirtualization2021}~\cite{zhangEvaluatingOptimizingVirtualization2010}. This is not desired and can be particularly problematic for real-time systems that require strict timing constraints. Making sure that tasks run predictably and on time in a virtualized environment, can be challenging. The hypervisor and other virtual components can cause variations in how long tasks take to execute~\cite{garcia-vallsChallengesRealtimeVirtualization2014}. When there are multiple virtual machines (VM) sharing the same physical resources, the VMs can affect each other's performance and consequently violate real-time constraints~\cite{queirozContainerbasedVirtualizationRealtime2024}.

@@

\subsection{Memory Management}
\noindent For the sake of completeness, Figure~\ref{fig:memory_management} displays the memory management of Salamander 4. LRT stands for Lasal Runtime and creates an execution environment where applications developed using LASAL Class 2 can run- It provides real-time functions, data types, and other constructs for real-time programming.

@@

\noindent The inital QEMU script after the custom Yocto build and the starting point for this work is shown in Code~\ref{script:qemu_def}. This script is used to start QEMU with required parameters to boot Salamamder 4 OS. It will be adjusted in Chapter~\ref{sec:real-time_tunings} in order to accompany real-time performance tunings.

>>>>

\hypersetup{
    colorlinks=false, % Aktiviert farbige Links
    linkcolor={rgb,255:red,0;green,0;blue,255}, % Interne Links (z.B. Sections)
    citecolor=orange, % Zitate
    urlcolor={rgb,255:red,128;green,0;blue,128} % URLs
}

@@

\bigskip \noindent Erste Latenzmessungen zeigten eine signifikante Leistungslücke zwischen der Bare-Metal und dem virtualisierten Setup. Daher wird ein umfangreicher Tuning-Prozess durchgeführt, um Echtzeit-Performance und Determinismus zu erreichen. Diese Modifikationen umfassen Konfigurationen, die das BIOS, den Kernel, das Host-Betriebssystem, die QEMU/KVM-Virtualisierungsschicht und das Salamander 4-Betriebssystem (Gast) selbst betreffen. Die Worst-Case-Latenz wurde von 707.622~µs auf 17.134~µs reduziert, was der Bare-Metal-Performance von 10.709~µs nahekommt. Darüber hinaus wird die Verbesserung der Echtzeit-Performance und des Determinismus anhand einer Roboteranwendung validiert, bei der die abgestimmte Virtualisierung mit der nicht abgestimmten und der Bare-Metal-Version verglichen wird.

\bigskip \noindent Insgesamt bietet diese Arbeit eine umfassende Methodik, um ein virtualisiertes Gastsystem in einem Hostsystem mit deterministischem Verhalten echtzeitfähig zu machen.}
\schlagworte{Virtualisierung, Echtzeitsysteme, Latenzreduktion, Robotiksteuerung}

@@

\bigskip \noindent Initial latency measurements revealed a significant performance gap between the bare metal and virtualized setup. Thus, an extensive tuning process is carried out to achieve real-time performance and determinism. These modifications involve configurations spanning the BIOS, kernel, host operating system, QEMU/KVM virtualization layer, and the Salamander 4 operating system (guest) itself. The worst-case latency was brought down from 707.622~µs to 17.134~µs, closely aligning with the bare metal performance of 10.709~µs. In addition, the improvement in real-time performance and determinism is validated using a robotic application, where the tuned virtualization is compared with the untuned and the bare metal version.

@@

\noindent However, in an RTOS, a high-priority process can interrupt a lower-priority task, even if it is in the middle of a kernel operation, as shown in Figure~\ref{fig:kernel_rt}. An RTOS is specifically designed to react to events within fixed time limits and prioritize the execution of high-priority processes.

@@

\section{Real-Time Operating Systems}
\noindent The RTOS structure is visualized in Figure~\ref{fig:rtos_structure}. The core component of an RTOS that enables real-time capabilities is the kernel. It is responsible for managing system resources, scheduling tasks, and ensuring deterministic behavior~\cite{malallahComprehensiveStudyKernel2021}. It employs preemptive scheduling mechanisms to allow high-priority tasks to preempt lower-priority tasks, ensuring that time-critical tasks are not delayed. Additionally, RTOS kernels are designed to allocate and manage memory resources efficiently and minimize interrupt latency, which is crucial for real-time applications that require immediate response to external events~\cite{wangRealtimeEmbeddedSystems2017}.

@@

\noindent In these real-time operating systems, task scheduling is based on so-called priority-based preemptive scheduling~\cite{buttazzoHardRealtimeComputing2024}. Each task in a software application is assigned a priority. A higher priority means that a faster response is required. Preemptive task scheduling ensures a very fast response. Preemptive means that the scheduler can stop a currently running task at any point if it recognizes that another task needs to be executed immediately. The basic rule on which priority-based preemptive scheduling is based is that the task with the highest priority that is ready to run is always the task that must be executed. So if both a task with a lower priority and a task with a higher priority are ready to run, the scheduler ensures that the task with the higher priority runs first. The lower priority task is only executed once the higher priority task has been processed.

@@

\begin{itemize}
	\item This work was written at SIGMATEK GmbH \& Co KG, a company that researches, develops, produces, and distributes automation technology for industrial machinery and plant engineering. SIGMATEK uses its own custom Linux-based operating system, to be run on their self-manufactured CPUs. This operating system will be referred to as ``Salamander 4'' in this work. The details of Salamander 4 are explained in Chapter~\ref{sec:guest_operating_system}.
	\item Salamander 4 was created with Yocto~\cite{WelcomeYoctoProject}, an open-source project that helps develop custom Linux-based systems. Yocto is discussed in Section~\ref{sec:yocto}.
	\item Salamander 4 is virtualized through Quick emulator (QEMU), an open-source emulator and virtualizer that allows running different operating systems on a computer. QEMU is described in Section~\ref{sec:qemu}.
	\item Salamander 4 employs hard real-time with Xenomai 3~\cite{XenomaiXenomai}, a real-time framework for Linux that enables deterministic and low-latency performance. Xenomai 3 is detailed in Section~\ref{sec:xenomai}.
	\item The goal is to virtualize Salamander 4 and approach the performance of bare metal (hardware) through real-time performance tunings, focusing on latency requirements set by SIGMATEK. These modifications are the subject of Chapter~\ref{sec:real-time_tunings}.
	\item The \texttt{latency} tool of the Xenomai suite to measure the scheduling latency of the real-time OS was suggested to be used by SIGMATEK. The goal was set to achieve latency values below 50 microseconds in the duration of the measurement.
	\item SIGMATEK uses the VARAN bus for communication between machines and systems, a real-time Ethernet network that was specially developed for industrial automation. Its operating principle is outlined in Section~\ref{sec:varan}.
	\item For the purpose of replacing the physical CPU, SIGMATEK provided a PCI Insert Card module PCV 522, that was plugged into the PC, as explained in Subsection~\ref{subsec:setup_experiment_virtualized}. This component could then be used to communicate with the Input/Output peripherals, just like the physical CPU could when the I/O modules were attached to it. Subsequently, the PCV 522 module of SIGMATEK was connected with the Pulse Width Module PW 022 of SIGMATEK over the VARAN bus and the VI 021 module of SIGMATEK serving as the power supply.
	\item The robotic application from Chapter~\ref{sec:robotic_application} was written in Lasal Class 2, an object-oriented programming tool of SIGMATEK and part of SIGMATEK's Engineering Platform LASAL, with client-server technology and graphic representation.
\end{itemize}

\section{Related Work and State of the Art}
This master's thesis builds upon a number of previous studies in the field of virtualization of real-time systems and the inherent latency. In this section, the most influential studies are presented and their relevance to the work is addressed.

\bigskip \noindent \citeauthor{perneelRealtimeCapabilitiesStandard2015}~\cite{perneelRealtimeCapabilitiesStandard2015} mention in their work that Linux was initially developed as a general-purpose operating system without real-time applications in mind. They talk about the evolution of real-time behavior in the standard Linux kernel and state that it has recently gained popularity among the real-time community, largely due to its open-source nature and stability. The authors explain how to transform Linux into a real-time operating system by implementing the PREEMPT-RT patches. The paper also emphasizes some kernel configurations that ensure that the system's real-time behavior is preserved during runtime. Even though these modifications to the kernel enable soft real-time performance, the authors underline the fundamental rule in real-time software, that latency improvements have a negative impact on throughput and kernel performance. This is due to the added overhead of the \texttt{CONFIG\_PREEMPT\_RT} option. They concluded with their Linux build and a testing application that Linux can be a viable option for real-time applications, if it is correctly configured. \citeauthor{reghenzaniRealTimeLinuxKernel2020}~\cite{reghenzaniRealTimeLinuxKernel2020} provide an extensive overview of the real-time Linux kernel research, specifically on the PREEMPT RT patch. \citeauthor{adamRealTimePerformanceResponse2021}~\cite{adamRealTimePerformanceResponse2021} applied the PREEMPT\_RT patch on both Raspberry Pi 3 and BeagleBone Black and achieved significantly lower latencies compared to standard Linux kernels.

\bigskip \noindent \citeauthor{cinqueVirtualizingMixedCriticalitySystems2022}~\cite{cinqueVirtualizingMixedCriticalitySystems2022} survey different virtualization approaches, including separation kernels, microkernels, enhanced general-purpose hypervisors, and lightweight virtualization solutions like containers and unikernels. Similarly, \citeauthor{sandstromVirtualizationTechnologiesEmbedded2013}~\cite{sandstromVirtualizationTechnologiesEmbedded2013}, and~\citeauthor{taccariEmbeddedRealTimeVirtualization2014}~\cite{taccariEmbeddedRealTimeVirtualization2014} discuss and compare the current state-of-the-art in virtualization technologies with a focus on embedded real-time platforms. The latter emphasizes that research on real-time systems is focusing on virtualization and multicore scheduling and underlines that flexibility and cost are key metrics. Both works acknowledge the great benefits of virtualization, that is, reducing overall hardware costs, since all non-critical activities and real-time control tasks can run on the same hardware. In addition,~\citeauthor{javierperezHowRealTime2022}~\cite{javierperezHowRealTime2022} state that hardware-based Programmable Logic Controllers (PLCs) are costly to maintain and lack the flexibility needed for modern, resource-intensive applications like Machine Learning (ML) or Artificial Intelligence (AI).

@@

\bigskip \noindent \citeauthor{maPerformanceTuningKVMbased2013}~\cite{maPerformanceTuningKVMbased2013}~\cite{junzhangPerformanceAnalysisKVMBased2010} describe in their work how KVM impacts real-time performance, which has helped this work gain insight into the base overhead of virtualization. When a physical interrupt occurs while the guest RTOS is running, there are at least six stages that need to be traversed before the interrupt can be delivered to the guest RTOS. Once the interrupt is delivered to the guest RTOS, it incurs additional latency, called Interrupt Routine Time (IRT), which is the time between when an interrupt is recognized and the first instruction of the corresponding interrupt service routine (ISR) of the guest OS is started. These stages are the primary sources of latencies caused by KVM, they are briefly described in Figure~\ref{fig:kvm_stages}.

\begin{figure}[H]
	\centering
	\begin{tikzpicture}[node distance = 0.35cm, auto, % Adjusted node distance
			block/.style={
					rectangle,
					draw,
					fill=blue!20,
					text width=30em,
					rounded corners,
					align=center,
					inner sep=0.5em
				},
			line/.style={draw, -latex'}
		]
		% Place nodes
		\node [block] (init) {\scriptsize \textbf{Guest Code}\\  The guest OS is executing its normal operations.};
		\node [block, below=of init] (vmexit) {\scriptsize \textbf{VM Exit}\\  External interrupt triggers VM exit. Hardware stores interrupt info in \\\vspace{-0.5em} Virtual Machine Control Structure (VMCS).};
		\node [block, below=of vmexit] (aftervmexit) {\scriptsize \textbf{After VM Exit}\\  Control returns to kernel space. KVM retrieves interrupt info from VMCS,\\\vspace{-0.5em} marks it as pending for guest.};
		\node [block, below=of aftervmexit] (hostisr) {\scriptsize \textbf{Host ISR}\\  Host Linux handles the interrupt.};
		\node [block, below=of hostisr] (hostscheduling) {\scriptsize \textbf{Host Scheduling}\\  Host kernel invokes scheduler. Ideally, guest RTOS process is not preempted.};
		\node [block, below=of hostscheduling] (beforevmentry) {\scriptsize \textbf{Before VM Entry}\\  KVM's kernel module prepares for VM entry. Detects pending interrupt,\\\vspace{-0.5em} injects virtual interrupt into guest by writing to VMCS.};
		\node [block, below=of beforevmentry] (vmentry) {\scriptsize \textbf{VM Entry}\\  Hardware emulates conventional interrupt handling for guest based on VMCS info.};
		\node [block, below=of vmentry] (guestirt) {\scriptsize \textbf{Guest IRT}\\  Guest's interrupt handler is invoked to service the interrupt.};
		% Draw edges
		\path [line] (init) -- (vmexit);
		\path [line] (vmexit) -- (aftervmexit);
		\path [line] (aftervmexit) -- (hostisr);
		\path [line] (hostisr) -- (hostscheduling);
		\path [line] (hostscheduling) -- (beforevmentry);
		\path [line] (beforevmentry) -- (vmentry);
		\path [line] (vmentry) -- (guestirt);
	\end{tikzpicture}
	\caption[Flowchart of Operations in a virtualized Environment]{Flowchart of Operations in a virtualized Environment (Own Figure, inspired by~\citeauthor{maPerformanceTuningKVMbased2013})}
	\label{fig:kvm_stages}
\end{figure}
\vspace{1em}

@@

\bigskip \noindent \citeauthor{broskyShieldedProcessorsGuaranteeing2003}~\cite{broskyShieldedProcessorsGuaranteeing2003} report in their work about the shielded CPU model for enhancing real-time performance in symmetric multiprocessor systems. By dedicating one or more CPUs to high-priority processes and interrupts, the goal is to provide deterministic execution of real-time applications and interrupt responses. The authors mention that there is no user interface for setting process CPU affinity in the standard Linux, while there is one for setting interrupt CPU affinity via the \texttt{/proc/irq/*/smp\_affinity} files. These files were used in this master's thesis. According to the authors, the shielded CPU model is especially useful for tasks that require guaranteed interrupt response times, very fast interrupt responses, and high-frequency, deterministic execution.

@@

\bigskip \noindent \citeauthor{yoonRealTimePerformanceAnalysis2009}~\cite{yoonRealTimePerformanceAnalysis2009} use some of these real-time tunings in their research, including CPU shielding, memory locking, and spinning nanosleep, for controlling humanoid robots equipped with around 60 servo motors and sensors. They implement EtherCAT for real-time communication of distributed devices, whereas this thesis uses Varan. The authors highlight the importance of timely data processing in robots that collect data from their environment and respond accordingly through their sensors and actuators. The authors conclude with an experiment that their robotic system could achieve deterministic control of actuators and sensors even under heavy system load through the use of Linux with the real-time preemption patch. Their research has similarities to this master's thesis in terms of real-time tunings for robot control. However, this thesis also includes the virtualization aspect that is central.

\bigskip \noindent \citeauthor{kiszkaLinuxRealTimeHypervisor2009}~\cite{kiszkaLinuxRealTimeHypervisor2009} specifically focuses on Linux as a hypervisor and analyzes its real-time capabilities when using KVM and QEMU. Most importantly for this master's thesis, the author warns of starvation when the priorities of QEMU's threads are raised and lifted into a real-time scheduling class. Starvation is when a process does not get the resources it needs for a long time because the resources are being allocated to other processes. A way to reduce this risk is by not using more virtual CPUs (vCPUs) than there are actual processor cores. Moreover, Linux has a feature that can limit how much CPU time all real-time tasks can use in the host system, which was important information for the real-time tunings of this thesis later.

@@

\bigskip \noindent \citeauthor{mckenneyRealTimeVs2008}~\cite{mckenneyRealTimeVs2008} addresses the overheads associated with real-time Linux when using KVM and QEMU. The author lists several sources of overhead, including memory locking to avoid page-fault latencies, increased overhead of locking and interrupts due to more-aggressive preemption, threaded interrupts that permit long-running interrupt handlers to be preempted by high-priority real-time processes, real-time task scheduling that requires global scheduling, high-resolution timers for tens-of-microseconds accuracy and precision, and preemptible RCU which has slightly higher read-side overhead than Classic RCU. This information helped understand the trade-offs involved in real-time Linux systems.

\bigskip \noindent \citeauthor{deoliveiraDemystifyingRealTimeLinux2020}~\cite{deoliveiraDemystifyingRealTimeLinux2020} critique the traditional \texttt{cyclictest} tool used for measuring scheduling latency, highlighting its limitations due to its black-box approach and lack of theoretical grounding. This contributed to the decision not to use this tool to measure the scheduling latency in this work. Instead, the \texttt{latency} tool was used, as was determined in Chapter~\ref{sec:application_context}.

\section{Problem and Task Definition}
As stated previously, virtualization of real-time systems has many advantages in comparison to traditional hardware-based approaches. Solutions based on pure hardware are often built for a distinct purpose and are hence limited in flexibility. It is difficult to adapt the hardware topology to new requirements or new applications. Moreover, during periods when the demand is lower, expensive hardware may sit idle, leading to inefficient use of resources and a poor return on investment~\cite{gabrielResourceAwareParameterTuning2019}. Another crucial factor is, of course, the high cost factor associated with procuring and maintaining separate hardware for each real-time system~\cite{bhardwajVirtualizationCloudComputing2021}. Especially when organizations scale their operations, they have to add more physical machines, and the expenses rise rapidly~\cite{abbasiExploringOpenStackScalable2023}. Besides, one needs physical access to the hardware for any updates or maintenance, which is hard to achieve, especially for systems in remote or difficult-to-access locations. This can lead to system downtime and may not only disrupt the operations but can also result in lost productivity and potential revenue~\cite{mahfoudRealTimePredictiveMaintenanceBased2024}.

\bigskip \noindent Virtualization promises to overcome these challenges~\cite{queirozContainerbasedVirtualizationRealtime2024}~\cite{bhardwajVirtualizationCloudComputing2021}~\cite{cinqueEvaluatingVirtualizationFog2023}. However, virtualization also has its own challenges and pitfalls. The additional layers of abstraction can lead to increased overhead and latency~\cite{casiniLatencyAnalysisVirtualization2021}~\cite{zhangEvaluatingOptimizingVirtualization2010}. This is not desired and can be particularly problematic for real-time systems that require strict timing constraints. Making sure that tasks run predictably and on time in a virtualized environment can be challenging. The hypervisor and other virtual components can cause variations in how long tasks take to execute~\cite{garcia-vallsChallengesRealtimeVirtualization2014}. When there are multiple virtual machines (VM) sharing the same physical resources, the VMs can affect each other's performance and consequently violate real-time constraints~\cite{queirozContainerbasedVirtualizationRealtime2024}.

@@

\subsection{Memory Management}
\noindent For the sake of completeness, Figure~\ref{fig:memory_management} displays the memory management of Salamander 4. LRT stands for Lasal Runtime and creates an execution environment where applications developed using LASAL Class 2 can run. It provides real-time functions, data types, and other constructs for real-time programming.

@@

\noindent The initial QEMU script after the custom Yocto build and the starting point for this work is shown in Code~\ref{script:qemu_def}. This script is used to start QEMU with required parameters to boot Salamander 4 OS. It will be adjusted in Chapter~\ref{sec:real-time_tunings} in order to accompany real-time performance tunings.
